---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stratigraphr <img src="man/figures/logo.svg" align="right" style="float: right; height: 180px; margin-left: 1em">

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![CRAN status](https://www.r-pkg.org/badges/version/stratigraphr)](https://CRAN.R-project.org/package=stratigraphr)
[![R-CMD-check](https://github.com/joeroe/stratigraphr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/joeroe/stratigraphr/actions/workflows/R-CMD-check.yaml)
[![Test coverage](https://codecov.io/gh/joeroe/stratigraphr/graph/badge.svg)](https://app.codecov.io/gh/joeroe/stratigraphr)
<!-- badges: end -->

stratigraphr is a tidy framework for working with archaeological stratigraphy and chronology in R.
It includes tools for reading, analysing, and visualising stratigraphies (Harris matrices) and sequences as directed graphs;
helper functions for using radiocarbon dates in a tidy data analysis; 
and an R interface to OxCal's Chronological Query Language (CQL).

## Installation

You can install the development version of stratigraphr from GitHub with [pak](https://pak.r-lib.org/):

``` r
# install.packages("pak")
devtools::install_github("joeroe/stratigraphr")
```

## Usage

`stratigraph()` uses a data frame of stratigraphic relations to construct a *stratigraphic graph*:

```{r eg-stratigraph}
library(stratigraphr)

# Example stratigraphy from Harris (1979), figure 12:
harris12

stratigraph(harris12, "context", "above")
```

stratigraph objects are built on top of [tidygraph](https://tidygraph.data-imaginist.com/), which gives access to a powerful tidy interface for manipulating and analysing the stratigraphy.
It also works seamlessly with [ggraph](https://ggraph.data-imaginist.com/).
For example, to approximate a conventional 'Harris matrix' visualisation with ggraph:

```{r eg-ggraph}
library("ggraph")

# Example data after Harris 1979, Fig. 12
harris12 |>
  stratigraph("context", "above") |>
  ggraph(layout = "sugiyama") +
    geom_edge_elbow() +
    geom_node_label(aes(label = context), label.r = unit(0, "mm")) +
    theme_graph()
```

For further information on:

* **Graph-bases stratigraphic analysis**, see `vignette("stratigraphr")`.
* **Chronological query language**, see `vignette("cql")`.

This package previous contained functions for tidy radiocarbon data.
These were deprecated in v0.3.0 and moved to the [c14](https://c14.joeroe.io) package.
See [c14's introductory vignette](https://c14.joeroe.io/articles/c14.html) for further information.
